Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
Working on new barrier synchronization chapter

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/hello8.png}
\end{center}
\begin{code}
<{:input:../code/hello8.hny}>
\end{code}
\caption{\harmonylink{code/hello8.hny} Less interleaving}
\label{fig:hello8}
\end{figure}

@@

<{:const:Barrier}>

@@

\begin{figure}
\begin{code}
<{:input:../code/qbarrier.hny}>
\end{code}
\caption{\harmonylink{code/qbarrier.hny} Barrier synchronization with queues}
\label{fig:barrier}
\end{figure}

Barrier synchronization is a problem that comes up in high-performance
parallel computing.
It is used, among others, for scalable simulation.
A barrier is almost the opposite of a critical section:
the intention is to get a group of threads to run some code at the
same time, instead of having them execute it one at a time.
More precisely, with barrier synchronization the threads execute in rounds.
Between each round there is a so-called \emph{barrier} where threads wait
until all threads have completed the previous round, before they start the
next one.
For example, in an iterative matrix algorithm, the matrix may be
cut up into fragments.  During a round, the threads run concurrently,
one for each fragment.  The next round is not allowed to start
until all threads have completed processing their fragment.

\begin{comment}
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
\end{center}
\caption{High-level state diagram specification of barrier synchronization
with three threads and two rounds.  The three numbers specify how many
rounds each thread has completed.}
\label{fig:barrierdiagram}
\end{figure}

\autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
with three and threads.  Initially all threads are in round~0.  Then each thread
can start and finish the round.  However, none of the threads can
progress to the next round until all threads have finished the
current round.
\end{comment}

Blocking queues work well for implementing barrier synchronization.
\autoref{fig:barrier} shows an example.  There is a queue
for each of the <{N}> threads.
Before thread <{i}> enters a round, it first sends a message to every
other thread and then waits until it receives a message from
every other thread.
In this simple case, each message contains <{None}>, but in practice useful
information may be exchanged between the threads.

The <{round}> array is kept to check the correctness of this
approach.  Each thread increments its entry every time it enters
a round.  If the algorithm is correct, it can never be that two threads
are more than one round apart.

\begin{figure}
\begin{code}
<{:input:../code/barrier.hny}>
\end{code}
\caption{\harmonylink{code/barrier.hny} Generalized barrier synchronization}
\label{fig:genbarrier}
\end{figure}

More generally, barrier synchronization can be abstracted as follows.
We want to create a \texttt{Barrier}$(n)$ object, with operations
\texttt{enter()} and \texttt{exit()}.  It is helpful to use
a roller coaster car with $n$ seats as a metaphor:
\begin{itemize}
\item the car cannot contain more than $n$ people;
\item the car won't take off until $n$ people are in the car;
\item no new people can enter the car until all $n$ people have left it.
\end{itemize}
Notice there are two different waiting conditions:
\begin{enumerate}
\item waiting for the car to empty out;
\item waiting for the car to fill up.
\end{enumerate}

But this poses a complication.  Suppose, for example, that there are two
seats in the car, and there is one person in the car.  Does that mean that
the car is not yet full, or not yet empty?  We have to distinguish
those situations.  To this end, it is useful to think of the car as going through
$2 \cdot n$ stages:
\begin{itemize}
\item[] Stage 0: the car is empty;
\item[] Stage $1 ... n-1$: the car is filling up;
\item[] Stage $n$: the car is full;
\item[] Stage $n+1 ... 2n - 1$: the car is emptying out.
\end{itemize}

\autoref{fig:genbarrier} shows code that implements barrier synchronization
using these stages.
Method~\texttt{enter}() has two <{wait}> loops, one for each
waiting condition.
This is sometimes called \emph{double turnstile}.
Each loop uses a different condition variable.
The first loop waits until the car has emptied out, while
the second waits for the car to fill up.
\autoref{fig:barriertest} is a test program.  The threads check
that all threads within the barrier are in the same round.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/hello8.png}
\end{center}
\begin{code}
<{:input:../code/hello8.hny}>
\end{code}
\caption{\harmonylink{code/hello8.hny} Less interleaving}
\label{fig:hello8}
\end{figure}

@@

<{:const:Barrier}>
<{:const:bwait}>

\begin{figure}
\begin{code}
<{:input:../code/barrier.hny}>
\end{code}
\caption{\harmonylink{code/barrier.hny} Barrier implementation}
\label{fig:genbarrier}
\end{figure}

\begin{figure}
\begin{code}
<{:input:../code/barriertest2.hny}>
\end{code}
\caption{\harmonylink{code/barriertest2.hny} Demonstrating the double-barrier pattern}
\label{fig:barriertest2}
\end{figure}

@@

Barrier synchronization is a problem that comes up in high-performance
parallel computing.
The Harmony model checker uses it.
A barrier is almost the opposite of a critical section:
the intention is to get a group of threads to run some code at the
same time, instead of having them execute it one at a time.
More precisely, with barrier synchronization, the threads execute in rounds.
Between each round, there is a so-called \emph{barrier} where threads wait
until all threads have completed the previous round and reached the
barrier---before they start the next round.
For example, in an iterative matrix algorithm, the matrix may be
cut up into fragments.  During a round, the threads run concurrently,
one for each fragment.  The next round is not allowed to start
until all threads have completed processing their fragment.

A barrier is used as follows:
\begin{itemize}
\item <{b = Barrier(n)}>: initialize a barrier <{b}> for a collection of
<{n}> threads;
\item <{bwait(?b)}>: wait until all threads have reached the barrier
\end{itemize}

\autoref{fig:barriertest} is a test program for barriers.
It uses an integer array <{round}> with one entry per thread.
Each thread, in a loop, waits for all threads to get to the barrier
before incrementing its round number.  If the barrier works as
advertised, two threads should never be more than one round apart.

When implementing a barrier, a complication to worry about is that
a barrier can be used over and over again.  If this were not the
case, then a solution based on a lock, a condition variable,
and a counter initialized to the number of threads could be used.
The threads would decrement the counter and wait on the condition
variable until the counter reaches 0.

\autoref{fig:genbarrier} shows how one might implement a reusable barrier.
Besides a counter <{.left}> that counts how many threads still have to
reach the barrier, it uses a counter <{.cycle}> that is incremented
after each use of the barrier---to deal with the complication above.
The last thread that reaches the barrier restores
<{.left}> to the number of threads
(<{.required}>) and increments the cycle counter.
The other threads are waiting for the cycle counter to be incremented.
The cycle counter is allowed to wrap around---in fact, a single bit
suffices for the counter.

A common design pattern with barriers in parallel programs,
demonstrated in \autoref{fig:barriertest2}, is to use the barrier
twice in each round.  Before a round starts, one of the threads---let's
call it the coordinator---sets up the work that needs to be done
while the other threads wait.  Then all threads do the work and go
on until they reach a second barrier.  The second barrier is used
so the coordinator can wait for all threads to be done before setting
up the work for the next round.
